``Interestingness'' in Mathematical Theory Exploration

### 2019-04-29 PhDThesis/background_haskell.tex
Bunch more changes to thesis

\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth, upquote=true]
module Email (Email(), at, render) where

@@

at :: String -> String -> Maybe Email
at "" h  = Nothing
at u  "" = Nothing
at u  h  = Just (E u h)
\end{lstlisting}

The \hs{Email} type guarantees that its elements have both a user part and a
host part (modulo divergence), but it does not provide any guarantees about
those parts. We also define the \hs{at} function, a so-called ``smart
constructor'', which has the additional guarantee that the \hs{Email}s it
returns contain non-empty \hs{String}s. By ommitting the \hs{E} constructor from
the export list on the first line \footnote{The syntax \hs{Email()} means we're
  exporting the \hs{Email} type, but not any of its constructors.}, the only way
\emph{other} modules can create an \hs{Email} is by using \hs{at}, which forces
the non-empty guarantee to hold globally.

@@

\item The existence of theory exploration systems such as \hspec{}, and related
  tools which we can to re-use, including conjecture generators like \qspec{}
  and \textsc{Speculate}; counterexample finders like \qcheck{}, \textsc{SmallCheck},
  \textsc{SmartCheck}, \textsc{LeanCheck}, \textsc{Hedgehog}, etc.; theorem
  provers like \textsc{Hip} \cite{rosen2012proving}; and other related testing
  and term-generating systems like \textsc{MuCheck} \cite{le2014mucheck},
  \textsc{MagicHaskeller}~\cite{katayama2011magichaskeller} and
  \textsc{Djinn}~\cite{augustsson2005djinn}.

@@

Whilst we will use normal Haskell code in this text and many of our examples,
for simplicity we perform many of our analyses on an intermediate representation
of the \textsc{GHC} compiler, known as \emph{GHC Core}, rather than the
relatively large and complex syntax of Haskell proper. Core is based on \fc{},
which is described in detail in~\cite[Appendix C]{sulzmann2007system}.

Core contains explicit type annotations and coercions, which we can omit as they
have no effect on runtime behaviour. The resulting sub-set of Core\footnote{As
  of GHC version 7.10.2, the latest at the time of writing.} is shown in
Figure~\ref{fig:coresyntax}; for brevity, we also omit several other forms of
literal (machine words of various sizes, individual characters, etc.), since
their treatment is similar to those of strings and numerals. We use quoted
strings to denote names and literals, e.g. \texttt{Local "foo"},
\texttt{Global "bar"}, \texttt{Constructor "Baz"}, \texttt{LitStr "quux"} and
\texttt{LitNum "42"}, and require only that they can be compared for equality.

@@

    even :: Nat -> Bool
    even    Z  = True
    even (S n) = odd n
  \end{haskell}
  \caption{A Haskell datatype for Peano numerals with some simple arithmetic
    functions, including mutually-recursive definitions for \hs{odd} and
    \hs{even}. \hs{Bool} is Haskell's built in boolean type, which can be
    regarded as \hs{data Bool \equal{} True | False}.}
  \label{fig:haskellexample}
\end{figure}

>>>>

  \begin{haskell}
    module Email (Email(), at, render) where

@@

    at :: String -> String -> Maybe Email
    at "" h  = Nothing
    at u  "" = Nothing
    at u  h  = Just (E u h)
  \end{haskell}

  This \hs{Email} type can be constructed in two ways: directly, via the \hs{E}
  constructor, which requires both a user part and a host part (both
  \hs{String}s, for simplicity); or indirectly, via the \hs{at} function, which
  will only call \hs{E} if both \hs{String}s are non-empty (a programming
  pattern known as a ``smart constructor''). The export list in the first line
  contains the \hs{at} function, but not the \hs{E} constructor~\footnote{The
    syntax \hs{Email()} means we're exporting the \hs{Email} type, but not any
    of its constructors.}, so any code using this module is only able to create
  \hs{Email} values with non-empty parts.

@@

\item The existence of theory exploration systems such as \hspec{}, \qspec{} and
  \textsc{Speculate}.

\item Related tooling we can re-use such as counterexample finders (\qcheck{},
  \textsc{SmallCheck}, \textsc{SmartCheck}, \textsc{LeanCheck},
  \textsc{Hedgehog}, etc.), theorem provers
  (e.g. \textsc{Hip}~\cite{rosen2012proving}), and other testing and
  term-generating systems like \textsc{MuCheck}~\cite{le2014mucheck},
  \textsc{MagicHaskeller}~\cite{katayama2011magichaskeller} and
  \textsc{Djinn}~\cite{augustsson2005djinn}.

@@

Whilst our systems and experiments use normal Haskell code, for simplicity we
perform some of our analyses on an intermediate representation of the
\textsc{GHC} compiler, known as \emph{GHC Core}, rather than the relatively
large and complex syntax of Haskell proper. Core is based on \fc{}, which is
described in detail in~\cite[Appendix C]{sulzmann2007system}.

Core contains explicit type annotations and coercions, which we omit as they
have no effect on runtime behaviour. The resulting sub-set of Core\footnote{As
  of GHC version 7.10.2.} is shown in Figure~\ref{fig:coresyntax}; for brevity,
we also omit several other forms of literal (machine words of various sizes,
individual characters, etc.), since their treatment is similar to those of
strings and numerals. We use quoted strings to denote names and literals,
e.g. \hs{Local "foo"}, \hs{Global "bar"}, \hs{Constructor "Baz"}, \hs{LitStr
  "quux"} and \hs{LitNum "42"}, and require only that they can be compared for
equality.

@@

    even :: Nat -> Bool
    even    Z  = True
    even (S n) = odd n
  \end{haskell}
  \caption{A Haskell datatype for Peano numerals with some simple arithmetic
    functions, including mutually-recursive definitions for \hs{odd} and
    \hs{even}. \hs{Bool} is Haskell's built in boolean type, which can be
    regarded as \hs{data Bool = True | False}.}
  \label{fig:haskellexample}
\end{figure}

### 2019-04-29 PhDThesis/background_quickcheck.tex
Bunch more changes to thesis

\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
factorial 0 = 1
factorial n = n * factorial (n-1)

fact_base      = factorial 0 == factorial 1
fact_increases = factorial 3 <= factorial 4
fact_div       = factorial 4 == factorial 5 `div` 5
\end{lstlisting}

@@

\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
factorial 0 = 1
factorial n = let nPos = abs n
               in nPos * factorial (nPos - 1)

fact_neg = factorial 1 == factorial (-1)
\end{lstlisting}

>>>>

\begin{haskell}
  factorial 0 = 1
  factorial n = n * factorial (n-1)

  fact_base      = factorial 0 == factorial 1
  fact_increases = factorial 3 <= factorial 4
  fact_div       = factorial 4 == factorial 5 `div` 5
\end{haskell}

@@

\begin{haskell}
factorial 0 = 1
factorial n = let nPos = abs n
               in nPos * factorial (nPos - 1)

fact_neg = factorial 1 == factorial (-1)
\end{haskell}

### 2019-04-29 PhDThesis/background_theoryexploration.tex
Bunch more changes to thesis

    even :: Nat -> Bool
    even    Z  = True
    even (S n) = odd n
  \end{minted}
  \caption{Haskell datatypes for booleans and natural numbers, followed by some
    simple function definitions (with type annotations).}
  \label{fig:haskellexample}
\end{figure}

\qspec{} (version 1) is written in the Haskell programming language, and works
with definitions which are also written in Haskell. A thorough description of
Haskell, including its suitability for theory exploration, can be found
in~\ref{sec:haskell}. For illustrative purposes, some simple Haskell definitions
are shown in Figure~\ref{fig:haskellexample}. The following procedure is used to
generate conjectures, which are both plausible (due to testing on many examples)
and potentially interesting (due to being mutually irreducible):

@@

As an example, we can consider a simple signature containing the expressions
from Figure \ref{fig:haskellexample}:

@@

\begin{figure}
  % To reproduce, run 'quickSpec nat' in haskell_example/src/QuickSpecExample.hs
  \begin{haskell}
                  plus a b = plus b a
                  plus a Z = a
         plus a (plus b c) = plus b (plus a c)
                  mult a b = mult b a
                  mult a Z = Z
         mult a (mult b c) = mult b (mult a c)
              plus a (S b) = S (plus a b)
              mult a (S b) = plus a (mult a b)
         mult a (plus b b) = mult b (plus a a)
                 odd (S a) = even a
            odd (plus a a) = odd Z
           odd (times a a) = odd a
                even (S a) = odd a
           even (plus a a) = even Z
          even (times a a) = even a
plus (mult a b) (mult a c) = mult a (plus b c)
  \end{haskell}
  \caption{Equations conjectured by \qspec{} for the functions in Figure
    \ref{fig:haskellexample}; after simplification.}
  \label{fig:qspecresult}
\end{figure}

>>>>

    even :: Nat -> Bool
    even    Z  = True
    even (S n) = odd n
  \end{minted}
  \caption{Haskell datatypes for booleans and natural numbers, followed by some
    simple function definitions (with type annotations).}
  \label{fig:haskellteexample}
\end{figure}

\qspec{} (version 1) is written in the Haskell programming language, and works
with definitions which are also written in Haskell. A thorough description of
Haskell, including its suitability for theory exploration, can be found
in~\ref{sec:haskell}. For illustrative purposes, some simple Haskell definitions
are shown in Figure~\ref{fig:haskellteexample}. The following procedure is used to
generate conjectures, which are both plausible (due to testing on many examples)
and potentially interesting (due to being mutually irreducible):

@@

As an example, we can consider a simple signature containing the expressions
from Figure \ref{fig:haskellteexample}:

@@

\begin{figure}
  % To reproduce, run 'quickSpec nat' in haskell_example/src/QuickSpecExample.hs
  \begin{haskell}
                  plus a b = plus b a
                  plus a Z = a
         plus a (plus b c) = plus b (plus a c)
                  mult a b = mult b a
                  mult a Z = Z
         mult a (mult b c) = mult b (mult a c)
              plus a (S b) = S (plus a b)
              mult a (S b) = plus a (mult a b)
         mult a (plus b b) = mult b (plus a a)
                 odd (S a) = even a
            odd (plus a a) = odd Z
           odd (times a a) = odd a
                even (S a) = odd a
           even (plus a a) = even Z
          even (times a a) = even a
plus (mult a b) (mult a c) = mult a (plus b c)
  \end{haskell}
  \caption{Equations conjectured by \qspec{} for the functions in Figure
    \ref{fig:haskellteexample}; after simplification.}
  \label{fig:qspecresult}
\end{figure}

### 2019-04-29 PhDThesis/litreview_premiseselection.tex
Bunch more changes to thesis

Recently, a variety of alternative algorithms have also been investigated,
dubbed \textsc{MaSH}, \textsc{SInE}, \textsc{BliStr}, \textsc{HOLyHammer},
\textsc{MoMM}, \textsc{SNoW}, \textsc{MPTP 0.2}, \textsc{MaLARea} and
\textsc{MaLARea SG1}. Each one combines an ``off the shelf'' machine learning
algorithm with a custom feature extraction approach, is trained using a corpus
of existing theorems and their proofs, and evaluated by trying to re-prove a
selection of unseen theorems from the corpus (i.e. given only the theorem
statement and possible lemmas, not the existing proof).

@@

The premise selection problem for theorem proving is related to the signature
selection problem for theory exploration, in that both are trying to pick a
small number of terms (depending on the speed of the subsequent algorithm and
how long the user is willing to wait) from a large set, to optimise the
outcome of that algorithm (e.g. finding a proof quickly; or finding many
interesting relationships quickly). The major difference is that premise
selection has a particular ``target'' theorem statement, against which to
compare each term (e.g. for the ``relevance'' of their sub-terms). This is not
the case for theory exploration, so we must consider the interactions of each
term with all of the others when deciding what to select.

>>>>

Recently, a variety of alternative algorithms have also been
investigated~\cite{kuhlwein2013mash,hoder2011sine,urban2013blistr},
%                  MaSH             SInE,         BliStr
dubbed , \textsc{SInE}, \textsc{BliStr}, \textsc{HOLyHammer},
\textsc{MoMM}, \textsc{SNoW}, \textsc{MPTP 0.2}, \textsc{MaLARea} and
\textsc{MaLARea SG1}. Each one combines an ``off the shelf'' machine learning
algorithm with a custom feature extraction approach, is trained using a corpus
of existing theorems and their proofs, and evaluated by trying to re-prove a
selection of unseen theorems from the corpus (i.e. given only the theorem
statement and possible lemmas, not the existing proof).

@@

The premise selection problem for theorem proving is related to the signature
selection problem for theory exploration, in that both are trying to pick a
small number of terms (depending on the speed of the subsequent algorithm and
how long the user is willing to wait) from a large set, to optimise the outcome
of that algorithm (e.g. finding a proof quickly; or finding many interesting
relationships quickly). The major difference is that premise selection has a
particular ``target'' theorem statement, against which to compare each
term. This is not the case for theory exploration, so we must consider the
interactions of each term with all of the others when deciding what to select.
